---
title: "Homework1"
author: "Tara Dolan"
date: "9/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load required packages
```{r, message=FALSE}
packages <-c("data.table","tidyverse")
lapply(packages,require,character.only = TRUE)
```

**1.Draw a graph of the population model variables, including relationships linking data, population variables, and parameters**


**2. Reparameterize the population dynamics model so that the numbers at time t are a function of the growth rate r and the population size in 1978**  

**3. Fit the population dynamics model to the available abundance data. You can assume that the observation error variance does not change over time and is an estimable parameter.**  
```{r}
#data
BCBW <-data.table(Year=c(1978,1980,1981,1982,1983,1985,1986,1987,1988,1993,2001),
                  Number=c(4765,3885,4467,7395,6573,5762,8917,5298,6928,8167,10545),
                  CV=c(0.305,0.343,0.273,0.281,0.345,0.253,0.215,0.327,0.12,0.071,0.128))  

#We are going to linearize the model, so create a log of Nt  
BCBW <- mutate(BCBW,lnNumber=log(Number))

#model inputs  
t <-BCBW$Year #time
lnN1978 <-BCBW$lnNumber[BCBW$Year=="1978"] #log abundance at time 1978
N1978 <-BCBW$Number[BCBW$Year=="1978"] #abundance at time 1978
Nt <-BCBW$Number # unlogged observations
lnNt <-BCBW$lnNumber # logged observations

#linear fit (unlogged)
fit_lm1 <-lm(Nt~t)
coef(fit_lm)
intercept <-fit_lm$coefficients[[1]]
growth_coeffcient <-fit_lm$coefficients[[2]]

#linear fit - not sure where the intercept comes in here. 
fit_lm <- lm(log(Nt) ~ t)
coef(fit_lm)
intercept <-exp(fit_lm$coefficients[[1]])
growth_coeffcient <-fit_lm$coefficients[[2]]

#nonlinear fit
add_nls <- nls(Nt ~ N1978*exp(r*t), start = list(r = 0.04))
coef(add_nls)
#doesn't work... maybe starting growth rate is bad? 


```

**4.Plot the resulting model predictions from the fitted model along with the data, and provide comment on the fit to the data (e.g. via residuals, etc.)**

```{r}

BCBW %>%
  ggplot(aes(Year,Number))+
  geom_point()
  
```  


